Versioned data: why it is needed and how it can be achieved (easily and cheaply)

### 2017-10-16 ms.tex
Edits up to discussion
Closes #29, closes #30, closes #24

\title{Versioned data: why it is needed and how it can be achieved (easily and cheaply)}
\author{Daniel S. Falster$^1,*$, Richard G. FitzJohn$^2$,\\ Matthew
  W. Pennell$^3$, and William K. Cornwell$^{1,*}$}
\affiliation{
$^1$ Evolution \& Ecology Research Centre, School of Biological, Earth and Environmental Sciences,\\
$^2$ School of Public Health, Imperial College, London SW7 2AZ United Kingdom\\
University of New South Wales, Sydney, NSW 2052, Australia\\
$^2$ Imperial College, London, United Kingdom\\
$^3$ Department of Zoology and Biodiversity Research Centre,\\
University of British Columbia, Vancouver, B.C. V6T 1Z4 Canada\\
$^*$ Corresponding author: daniel.falster@unsw.edu.au\\
}
\date{}

@@

% For Analyses and Articles, the main text (excluding abstract, Methods, references and figure legends) is approximately 3,000 words. The abstract is typically 100-170 words, unreferenced. An Introduction is followed by sections headed Results, Discussion, Methods. The Methods and Results should be divided by topical subheadings; the Discussion does not contain subheadings. The Methods should be followed by References, Data Citations (optional), Acknowledgements and a Competing interests statement. A Data Citation section should be included when externally hosted datasets are referred to in the work and each can be cited formally with a unique, stable identifier and repository name.

@@

The sharing and re-use of data has become a cornerstone of modern science.  To help this process, numerous linked tools have been developed to allow quick and easy data sharing:    a data file for a particular analysis can be quickly uploaded, given a digital identifier, and shared.  Thus far, however, this framework does not accommodate on-going scientific progress: for many important problems, particularly those suitable for meta-analyses, datasets will continue to grow with time --- more data points will be added, and new and more useful data structures will be created.  Essentially this means that datasets, like science in general, improve with time.  We suggest that many different datasets would be more useful if organized as a series of versions, with a simple naming system to allow users to perceive the type of change between versions.  In this article, we argue that adopting the paradigm and processes for versioned data, analogous to software versioning, would be useful for both data curators and users of even small-to-medium sized datasets. We introduce a system called Versioned Data Delivery (VDD) and present tools for creating, archiving, and distributing versioned data easily, quickly, and cheaply. These new tools allow for individual research groups to shift from a static model of data curation to a dynamic and versioned model that more naturally matches the scientific process.

@@

Many, and perhaps most, of the datasets published in \emph{Scientific Data} are likely to be living entities, meaning the state-of-the-art version of the dataset will change through time. Typical changes include: the adding new data, improving the quality of existing data, integrating with other databases, or re-structuring the database content to be able to address new questions. For example, a dataset on a biological organisms might be expanded through the addition of new records or improved through the correction of spelling mistakes in taxonomic names. As research around a data product grows, there might be many such additions. Ideally, any changes to a dataset would become immediately available to any interested parties, while past versions of the dataset would remain accessible to ensure reproducibility.

However, current models for publishing datasets do not facilitate distributing successive versions of a dataset in an effective and scalable manner. The most common current working model for publishing scientific datasets is to archive them in a variety of open-source data repositories, such as \texttt{Dryad}, \texttt{Figshare}, or \texttt{Zenodo}, where the data is attributed a Digital Object Identifier (\textsc{doi}). Once published, the datasets are immutable. That means that most published datasets exist as a singular snapshot of the dataset, taken at the time the data descriptor was published. While it might only require a small amount of work to update the dataset, if there is no way to distribute the updated dataset, then that work has to be re-done by every user of that dataset. Moreover, those required changes add steps between the canonical dataset and any analysis using it, making reproducibility of the new analysis more difficult than it needs to be.

Many research groups internally solve the issue of versioning data in a variety of suboptimal ways. A common solution is to email around the latest versions with version numbers appended to the filename. Another approach is to repeat corrections or updates that have been made elsewhere. This practice is both inefficient and a barrier to reproducibility, as the dataset used in a publication may differ to the version previously made available on-line. There is therefore a strong need for an easy way to distribute changes to a dataset after its initial publication to potential users, along with notes on what has changed and why since the previous version.

In this article we i) introduce the concept of Versioned Data Delivery (\textsc{vdd}); ii) outline how emerging technologies in data science (Table \ref{tab:technologies}) can be used to help researchers maintain, distribute, and access small-to-medium sized datasets; and iii) introduce a new \textsc{R} package called \texttt{datastorr}, a proof-of-concept implementation of a \textsc{vdd} system. The issue of updating and expanding published data has been partly addressed in large centralised repositories like genetic sequences (\texttt{GenBank}) or species location data (\texttt{GBIF}), where new data can be added and there exist abilities to correct errors in existing records (e.g. by adding multiple version of a genetic sequence). Yet, such large centralised repositories require a level of infrastructure that is beyond most research groups, or even countries (Table \ref{tab:sql_v_versioneddata}). Our focus here is on the wide range of datasets, such as those appearing in this journal, that are not covered by these repositories. In most of these cases, the data collected will not be ``big'' but rather small-to-medium sized. Such data support important research projects on particular scientific questions, but are not in most cases general enough to warrant custom infrastructure. Further, while we emphasize particular technologies in our implementation, the principles are general and could easily be ported to other platforms. Likewise, while the examples cited here come from our own research areas, the issues we have identified and the solutions we have put forward are relevant to nearly all data-oriented scientific domains.

% I think that this model is actually ok; for data archiving where the
% data are a record of what was used in the paper, a single immutable
% data set makes sense.  It just doesn't extend to things where the
% research output *is* the dataset.  So this is more a case of
% pointing out that this model does not apply well to all sorts of
% data - given how successful the archive-at-publication model has
% been at getting data out of people's drawers I don't think we should
% complain that it doesn't do something that it does not need to do.

% A particular issue with the static
% model is that data releases are mostly triggered by publication of academic
% papers. But for a living dataset, there may be multiple releases made in
% between papers. This is particularly the case with the discovery of small
% errors.
% % OK, here again; it's not so much medium-sized as *living*.  The
% % issues in practically doing the versioning are more acute when the
% % data are bigger than small, that's all.
% For any living dataset that is put to reuse, errors will
% inevitably be found, and there needs to be some easy  way to correct them and
% distribute the updated dataset.

% To overcome the limitations of static datasets, many research groups have
% setup dedicated webservers to host and distribute data (Table 1)\citep[e.g.][]{Madin-2016}. In these systems,
% datasets areas maintained in a structured database, such as SQL, and a custom front
% end to this database must be written.
% When designed well, such systems are very useful for updating or adding to a
% dataset. However, the infrastructure investment (designing maintaining, and
% hosting the database) is overkill for most projects and is a very
% substantial barrier to entry. Furthermore, the system requires ongoing funds to maintain it, making it often unsustainable for many research groups. Moreover, the existence of a live webserver,
% does not in itself guarantee ongoing access to different versions of a
% dataset.

@@

% I really like the focus on "release".  It's a happy coincidence that
% it's what github uses in the interface.

In brief, the workflow we present, which we call Versioned Data Delivery (\textsc{vdd}), borrows best-practices for software development \citep{Perez-Riverol-2016} and applies them to the challenge of maintaining and distributing versioned data. Software developers maintain a core set of code which produces a binary executable file that can be installed on a users local computer. With a \textsc{vdd} system, data developers similarly maintain a core set of files (the ``code''), which produce an organised dataset that can be ``installed'' on a user's local computer. In either the development of software or data, successive versions are released over time (called ``releases''). The similarity in workflow between software and data then allows us to deploy the same technological platforms that are used in software development, for the development and distribution of data. Importantly, our \textsc{vdd} systems uses well-established tools, ensuring high-level performance and stability.

The core technologies used are summarised in Table \ref{tab:technologies} and described further below.

\subsection{Version control and `semantic' versioning}

% You could
% talk about 'git' (referencing Ram-2013) and say "on the face of it,
% versioning tools like git appear to provide the ideal interface for
% versioning data".  Then talk about how git is designed for code and
% breaks down as the data involved exceed a few 10s to hundreds of MB.
% After that, we can say that the aim is not to support truely "Big"
% data because research groups capable of generating such data
% probably have the wherewithall to look after it.  I know you bring
% it up below, but I think it would go better here...

% With suggested edits, some of this moves up.

Version control, primarily an open-source variety called \texttt{git}, has become ubiquitous in software development. In practice, version control tracks line-by-line changes in text files and creates and maintains a history of those changes. Increasingly version control has been applied to scientific code and data management, especially for small-to-medium sized datasets \citep{Ram-2013, Perkel-2016}. \texttt{git} is attractive for data management because it tracks all changes in monitored files. This history is visible to anyone interacting with the repository. It also allows users to annotate changes (`commits') with informative messages detailing the rationale for those changes.

To fully realize the benefits of a versioned data, the \textsc{vdd} system needs to be transparent and understandable by both data sharers and users.  Crucially a user should be immediately aware of the relationship among versions.  Software development has dealt with a very similar problem, and we suggest there is benefit in adopting the notation from that field.  Specifically, we suggest applying the theory of ``semantic versioning'' developed for software distribution. This system uses (the potentially familiar) notation of form ``1.0.0'' for successive versions (Fig \ref{fig:semantic}). As has been noted (REFS), labelling of data versions presents a similar problem to the labelling of software releases: signalling to users the types of change that has occurred between successive versions. Applying semantic versioning to data, a change from 1.0.0 to 1.0.1 implies tiny changes, for example a minor error correction which is unlikely to ``break'', or change in a substancial way, users' analyses, although there may be minor changes in the results. A change from 1.0.0 to 1.1.0 implies a substantial enhancement, for example adding a new study to a meta-analysis dataset (while otherwise maintaining the same dataset structure); this change is large enough that in the view of the database curator it might change the results of some users' analyses. A change from 1.0.0 to 2.0.0 implies a very major revision, for example improving the entire structure of the dataset and adding new columns. These changes are very likely to change the results of most users' analyses or to break code that was written to work on the previous version of the database or both.  
% I think that major version increments could still be looked at as
% _breaking_ changes as often as large additions of data.  Changing
% column names, removing columns, changing the structure of linked
% tables etc.

\subsection{Distributing data to different user needs}

Currently there are two types of data users: those that  interact with the data via point and click downloading and those that use code-based interaction.  The system we describe allows for both types (Figure \ref{fig:technology_stack} and Table \ref{tab:user_requirements}).  \texttt{GitHub} has a feature called ``releases'' in which specific commits can be turned into a version for distribution (with associated version number).  Those releases can be downloaded directly by users via point-and-click, or accessed programmatically via the \texttt{GitHub} \textsc{api}.  This system also allows the curator to do internal work between releases and only \emph{release} the data to users when the revision represents a clear improvement on the previous release.

% I have no idea what this next sentence means.  WKC: cut
%Releases can also be embedded into any kind of website that allows the data curator to facilitate discovery by more users and to provide context.  Releases can also be linked to digital object identifiers and code-based users via for example an \textsc{R} package.

@@

A point-and-click workflow may create barriers for reproducibility \citep{Wilson-2014,Lowndes-2017}. Ideally, users should be able to access all versions of the database programmatically. Code to access the \texttt{GitHub} releases could be written individually by each user, but this creates an unnecessary technological hurdle for reproducibility.  To make access via code as easy as possible for users, we offer a novel implementation of a \textsc{vdd} system. We have written an \textsc{R} package, \texttt{datastorr} (\smurl{github.com/richfitz/datastorr}), that a) interacts with the \texttt{GitHub} \textsc{api} to pull down versions of the dataset; and b) constructs the shell of a second, database specific \textsc{R} package, which can be used to access versions of the dataset. Using this system, a researcher can create and distribute a \textsc{R} package that facilitates access to their data with (very) minimal computational skills. We have focussed on the \texttt{R} platform \cite{R-2017} as one of the most prominent platforms for data analysis.

For example, \texttt{datastorr} was used to build the package \texttt{baad.data} (\smurl{github.com/traitecoevo/baad.data}), which is an interface to the Biomass and Allometry Database (\textsc{BAAD}) \citep{Falster-2015} stored at \smurl{github.com/dfalster/baad}. \texttt{baad.data} consists of only a few simple functions, that were automatically generated (along with the associated help files) with \texttt{datastorr}. And importantly \texttt{baad.data} contains no actual data so it is very quick to install and takes up virtually no space on the user's hard-drive.
% Dan Noble - Next 5 lines: I found this part to be a bit difficult to grasp as you discuss how updated versions can "download the latest version”, but then in the same sentence discuss how this works seamlessly “offline”. You may want to discuss this in greater depth as I don’t see how this is possible.
When the function \texttt{baad.data::baad\_data()} is called, the system will download the latest version; if there have not been any changes since the last time the function was called, it will simply retrieve the data from memory, making the system very efficient (and also allowing the system to work seamlessly if the computer is offline). If a researcher requires an older version of \textsc{BAAD}, specifying the version number of the data as an argument, e.g. \texttt{baad.data::baad\_data("1.0.1")} will retrieve it, without overwriting any other dataset versions they may have previously downloaded. This enables users to reproduce an analysis using a specific version of the data, even while new data is posted.  Users may also have several versions of the same database on their computer and unambiguously access the different versions with one simple function call. The dataset versions are cached between runs such that the data are accessed by name and version, so that after the first download of each version access to the data should be very fast.

% Something not mentioned; that the versions are cached between runs
% so that the data are accessed by name/version (rather than by
% homebrewing a local cache and manually doing a bunch of file name
% mangling) and that all accesses after the first are very fast
% WKC: I had a go at this--see above

Using \texttt{datastorr}, researchers can set up their own versioned database (and R package) simply by providing: 1) a \texttt{GitHub} repository name (e.g., ``traitecoevo/taxonlookup''); 2) a single file that will contain the versioned data (e.g., ``taxonlookup.csv''); 3) and the function used to load this into \texttt{R} (e.g., \texttt{read.csv()}). A tutorial explaining precisely how to set this up is available at \smurl{github.com/ropenscilabs/datastorr}.

Then as the dataset grows over time, researchers update the \texttt{git} repo and create a \texttt{GitHub} release with a new version number (preferably
using \href{http://semver.org/}{semantic versioning}). All the releases are simultaneously
available to the user by both point-and-click and programmatically.

We have created \texttt{datastorr} to make \textsc{vdd} easy to implement and manage for \texttt{R} users. And as we stated above, this has already been used to release a variety of datasets (Table \ref{tab:examples}).
% Reword next senetnce
However, we can imagine alternative methods for interacting with \texttt{GitHub} and building templates and adopting a \textsc{vdd} system.  The key is not the specific technology, but rather the concept of creating, maintaining, and distributing versioned data.

\section{Discussion}
% General problem. An example of solution. Key features

@@

%work on this paragraph more
We strongly suspect that the issues we have identified in this article will not come as news to most readers: datasets are constantly evolving (hopefully improving) and despite tremendous advances in data sharing and associated technologies over the last decade, there is little consensus about what to do with this fact. In some individual cases, such as \texttt{GenBank} (see \url{https://www.ncbi.nlm.nih.gov/genbank/sequenceids/}), researchers have thought a great deal about data versioning and developed use-case specific solutions. But as we mentioned above, these datasets tend to be those managed by large, well-funded institutions or consortia and involve standard data types (e.g., sequences). But for an individual research group working with a specialized type of data, creating their own \texttt{GenBank} is clearly a non-starter. As a result, we argue that there is a substantial need for a easy, cheap, and flexible solution and suggest that adopting a \textsc{vdd} system may be one way forward.

Many of the key roadblocks preventing a switch from a static to a dynamic data world were technological: in the past, it took great deal of money and expertise to set up and maintain a \textsc{vdd} system, but now there are many current platforms for storing and distributing versioned controlled data for low or no cost.  The website \texttt{GitHub} (\smurl{www.github.com}) is a commercial platform for hosting and interacting with \texttt{git} repositories. In practice, each dataset should be a separate \texttt{git} repository on \texttt{GitHub} (see Table \ref{tab:examples} for examples). Although mainly used for computer code, \texttt{GitHub} is now also being use to manage scientific data \citep{Perkel-2016}. Maintaining the version controlled data in the cloud has two main benefits: first, it provides a platform for multiple data contributors to sync their files and correspond about changes in the dataset, and second, it hosts a stream of data releases, thus acting as a central point for  the collection, curation, and distribution of the data.  The proliferation of cloud-based tools and open-source software have now reduced these roadblocks considerably.  We argue that by linking together the tools described above, researchers can now produce, curate, and distribute versioned data much more easily and cheaply than in the past.

%The key features of a \textsc{vdd} system that are already possible are a versioning system to communicate small and large changes to users, both point-and-click and code-based access to all versions, reliable distribution to many users, and stable, organized \textsc{doi}s (see Figure \ref{fig:technology_stack}).  Moreover, it is relatively easy to update and add new versions.  Technology is moving forward rapidly and so it is very likely that the particular tools we describe above will add features and be even easier to use.  But, nonetheless, workable systems are already possible.

@@

%Systematically archiving data is a key part of good workflow \citep{Wilkinson-2016, Piwowar-2011, Whitlock-2011}, and tools for creating and archiving digital objects have seen rapid progress in recent years.  At present, each version of the dataset can be automatically archived and assigned a \textsc{doi}, via several providers (Table \ref{tab:doi_minting}).  Here we describe the particular integration of \texttt{GitHub} and \texttt{Zenodo}.  It is useful to have the ability to either be specific---reference a particular version for the purposes of reproducibility or general--reference the entire project.  \texttt{Zenodo} currently automatically supplies both ``concept'' and ``version'' \textsc{doi}'s to the commits associated with \texttt{GitHub} releases automatically \citep[Figure \ref{fig:semantic} and ][]{Nielsen-2017}.  This allows users to very precisely reference either generally or specifically, both of which may be appropriate in different situations.

@@

Because \texttt{git} was built for tracking changes in code, there are aspects of how it behaves that are not natural to apply to  data \cite{Perkel-2016}.  For example, because it tracks changes line-by-line rather than cell-by-cell, there are inefficiencies in how the software stores the changes.
There is rapid current work on new implementations which will be faster computationally, smaller in file size, flexible with respect to data structures, and thus allow for larger datasets to be versioned efficiently\citep{Fli, Dat}.
% Add CKAN, rdrop2 and Git-LFS, conceptually at least, and say something useful about the future

@@

>>>>

\title{Versioned data: why it is needed and how it can be achieved (easily and cheaply)}
\author{Daniel S. Falster$^{1,*}$, Richard G. FitzJohn$^2$,\\ Matthew
  W. Pennell$^3$, and William K. Cornwell$^{1}$}
\affiliation{
$^1$ Evolution \& Ecology Research Centre, School of Biological, Earth and Environmental Sciences,
University of New South Wales, Sydney, NSW 2052, Australia\\
$^2$ School of Public Health, Imperial College, London SW7 2AZ United Kingdom\\
$^3$ Department of Zoology and Biodiversity Research Centre,\\
University of British Columbia, Vancouver, B.C. V6T 1Z4 Canada\\
$^*$ Corresponding author: daniel.falster@unsw.edu.au\\
}
\date{}

@@

@@

The sharing and re-use of data has become a cornerstone of modern science.  To help this process, numerous linked tools have been developed to allow quick and easy data sharing:    a data file for a particular analysis can be quickly uploaded, given a digital identifier, and shared.  Thus far, however, this framework does not accommodate on-going scientific progress: for many important problems, particularly those suitable for meta-analyses, datasets will continue to grow with time --- more data points will be added, and new and more useful data structures will be created.  Essentially this means that datasets, like science in general, improve with time.  We suggest that many different datasets would be more useful if organized as a series of versions, with a simple naming system to allow users to perceive the type of change between versions.  In this article, we argue that adopting the paradigm and processes for versioned data, analogous to software versioning, would be useful for both data curators and users of even small-to-medium sized datasets. We introduce a system called Versioned Data Delivery (\textsc{vdd}) and present tools for creating, archiving, and distributing versioned data easily, quickly, and cheaply. These new tools allow for individual research groups to shift from a static model of data curation to a dynamic and versioned model that more naturally matches the scientific process.

@@

Many, perhaps most, of the datasets published in \emph{Scientific Data} are likely to be living entities, meaning the state-of-the-art (or ``canonical'') version of the dataset will change through time. Typical changes include: the adding new data, improving the quality of existing data, integrating with other databases, or re-structuring the database content to be able to address new questions. For example, a dataset on a biological organisms might be expanded through the addition of new records or improved through the correction of spelling mistakes in taxonomic names. As research around a data product grows, there might be many such additions. Ideally, any changes to a dataset would become immediately available to any interested parties, while past versions of the dataset would remain accessible to ensure reproducibility.

However, current models for publishing datasets do not facilitate distributing successive versions of a dataset in an effective and scalable manner. The most common current working model for publishing scientific datasets is to archive them in a variety of open-source data repositories where the data is attributed a Digital Object Identifier (\textsc{doi}) -- such as \texttt{Dryad}, \texttt{Figshare}, or \texttt{Zenodo}. Publishing involves a single data deposition. And once published, datasets are immutable. That means that most published datasets exist as a singular snapshot of the dataset, taken at the time the data descriptor was published. While it might only require a small amount of work to update the dataset, if there is no way to distribute the updated dataset, then that work has to be re-done by every user of that dataset. Moreover, those required changes add steps between the canonical dataset and any analysis using it, making reproducibility of the new analysis more difficult than it needs to be. We refer to this type of data delivery as being ``Static'' (Table \ref{tab:publishing_models}).

Many research groups solve the issue of versioning data internally in a variety of ad hoc and suboptimal ways. A common solution is to email around the latest versions with version numbers appended to the filename. Another approach is to repeat corrections or updates that have been made elsewhere. Alas, these practices are both inefficient and a barrier to reproducibility, as the dataset used in a future publications may differ to the version previously made available on-line. There is therefore a strong need for an easy way to distribute changes to a dataset after its initial publication to potential users, along with notes on what has changed and why since the previous version.

In this article we
\begin{enumerate}
  \item Introduce the concept of Versioned Data Delivery (\textsc{vdd}).
  \item Outline how emerging technologies in data science (Table \ref{tab:technologies}) can be used to help researchers maintain, distribute, and access small-to-medium sized datasets.
  \item Introduce a new \textsc{R} package called \texttt{datastorr}, a proof-of-concept implementation of a \textsc{vdd} system.
\end{enumerate}
The issue of updating and expanding published data has already been addressed in large centralised repositories like genetic sequences (\texttt{GenBank}) or species location data (\texttt{GBIF}), where new data can be added and there exist abilities to correct errors in existing records (e.g. by adding multiple version of a genetic sequence). Yet these ``Dynamic'' web databases (Table \ref{tab:publishing_models}), require a level of infrastructure that is beyond most research groups. Our focus here is on the wide range of datasets, such as those appearing in this journal, that are not covered by these repositories. In most of these cases, the data collected will not be ``big'' but rather small-to-medium sized. Such data support important research projects on particular scientific questions, but are not, in most cases, general enough to warrant custom infrastructure. Further, while we emphasize particular technologies in our implementation, the principles are general and could easily be ported to other platforms.

@@

In brief, the \textsc{vdd} workflow we present borrows best-practices for software development \citep{Perez-Riverol-2016} and applies them to the challenge of maintaining and distributing Versioned Data (\textsc{vd}). Software developers maintain a core set of code which produces a binary executable file that can be installed on a users local computer. With a \textsc{vdd} system, data developers similarly maintain a core set of files (the ``code''), which produce an organised dataset that can be ``installed'' on a user's local computer. In either the development of software or data, successive versions are released over time (called ``releases''). The similarity in workflow between software and data then allows us to deploy the same technological platforms that are used in software development, for the development and distribution of data. Importantly, our \textsc{vdd} systems uses well-established tools (Table \ref{tab:technologies}), ensuring high-level performance and stability.

The core technologies used are summarised in Table \ref{tab:technologies} and described further below. Several groups will interact with the \textsc{vdd} system, including dataset \emph{maintainers}, \emph{contributors}, and \emph{users}. The requirements of these different groups are outlined in Table \ref{tab:user_requirements}.

\subsection{Version control}

Version control, primarily an open-source variety called \texttt{git}, has become ubiquitous in software development. In practice, version control tracks line-by-line changes in text files and creates and maintains a history of those changes. Increasingly version control has been applied to scientific code and data management, especially for small-to-medium sized datasets \citep{Ram-2013, Perkel-2016, Lowndes-2017}. \texttt{git} is attractive for data management because it tracks all changes in monitored files, provided these are saved in text format (e.g. ``.csv'', ``.tsv'', ``.txt''). The history is visible to anyone interacting with the repository. It also allows users to annotate changes (``commits'') with informative messages detailing the rationale for those changes. In it's present form, \texttt{git} can handle individual data files at least up to 100MB, which includes a large fraction of scientific cases.

As a general strategy for tracking a dataset under version control with git, we recommend:
\begin{enumerate}
  \item Dataset \emph{maintainers} establish a separate \texttt{git} repository for any dataset to be distributed.
  \item Saving all files as plain text, so that \texttt{git} can identify line-by-line changes. For example, you might maintain tabular data as a ``csv''.
  \item Saving data in their rawest form. In some datasets you might only have a single file. Others may have may files that get manipulated or combined in some way to produce a unified product.
  \item Including in the \texttt{git} repository any code needed to manipulate or compile the raw data files into the final dataset. For example, you might combine many independent datasets into one unified dataset.
  \item Documenting any change in the dataset as a ``commit'' into the \texttt{git} repository logs, with informative commit messages outlining why a particular change was made.
\end{enumerate}

\subsection{Hosting and distributing versioned data}

Datasets stored under version control via \texttt{git} reach their real potential when hosted at a suitable internet hosting service \citep{Ram-2013,Perkel-2016}. Here we focus on the platform \texttt{GitHub} (Table \ref{tab:technologies}), but similar functions could be achieved via other providers such as \href{http://bitbucket.org}{bitbucket.org} and \href{http://gitlab.com}{gitlab.com}. Hosting of a \texttt{git} repository enables dataset \emph{maintainers} to connect with other potential \emph{contributors} and also \emph{users} (Fig. \ref{fig:technology_stack}). These platforms are designed to work with \texttt{git} repositories, and thus offer many helpful features, such as ability to record issues, host documentation, or review edits over time.

Of particular interest for current purposes, is \texttt{GitHub}'s ability to host a stream of ``releases'' from the dataset, alongside the \texttt{git} repository containing all the rawfiles. Each ``release'' is linked to a specific commit in the \texttt{git} repository history and occur at points where the dataset \emph{Maintainer} decided to generate a new  version of the data for distribution. While \emph{users} could in principle download the entire \texttt{git} repository, actually all they want are these 'releases'.

Deciding when to make a new release is at the discretion of the dataset \emph{Maintainer}. In practice, one makes fewer releases than one does commits into the \texttt{git} repository, though there is nothing stopping \emph{maintainers} from releasing a new version for every commit. The flexibility here allows \emph{maintainers} to do internal work between releases and only \emph{release} the data to users when the revision represents a clear improvement on the previous release.

Another important consideration is that websites like \texttt{GitHub} naturally cater for two types of data users to access the data: those that interact with the data via point and click downloading and those that use programmatic interaction (Fig. \ref{fig:technology_stack} and Table \ref{tab:user_requirements}). Specifically, \texttt{GitHub} releases can be downloaded directly by users via point-and-click, or accessed programmatically via the \texttt{GitHub} \textsc{api}.

\subsection{`Semantic' versioning}

To realize the full benefits of a versioned controlled dataset, users should be able to easily intuit the types of changes that have occurred among versions. Since software development has effectively already dealt with almost identical; problem in the labelling of software releases, we suggest there is benefit in adopting the best-practices from that field.

Specifically, we suggest applying the theory of ``semantic versioning'', developed for software distribution (see \href{http://semver.org/}{semver.org}), to successive releases of a dataset. Semantic versioning uses tri-digit notation of the form ``X.Y.Z'' for successive versions, where X, Y, and Z are non-negative integers. For example, version ``1.0.0''. Changes in the version number then signal the type of changes that occurred in the dataset (Fig \ref{fig:semantic}). Applying semantic versioning to data, a change from:
\begin{itemize}
  \item  {\bf 1.0.0 $\rightarrow$ 1.0.1}: implies a {\bf correction} or ``patch'', for example a small error correction which is unlikely to ``break'', or change in a substantial way, a users' analyses, although there may be minor changes in the results.
  \item {\bf 1.0.0 $\rightarrow$ 1.1.0}: implies a {\bf minor} enhancement, for example adding a new study to a meta-analysis dataset (while otherwise maintaining the same dataset structure); this change is large enough that in the view of the database curator it might change the results of some users' analyses.
  \item {\bf 1.0.0 $\rightarrow$ 2.0.0}: implies a very {\bf major} revision, for example improving the entire structure of the dataset and adding new columns. These changes are very likely to change the results of most users' analyses or to break code that was written to work on the previous version of the database or both.
\end{itemize}
Any user of a dataset where releases were tagged using semantic versioning, would immediately know the types of change that might have occurred when requesting different versions and set their there expectations accordingly.

% I think that major version increments could still be looked at as
% _breaking_ changes as often as large additions of data.  Changing
% column names, removing columns, changing the structure of linked
% tables etc.

@@

To aid reproducibility and efficient usage, many users will want access to all versions of any particular dataset programmatically (Table \ref{tab:user_requirements}). Code to access a stream of \texttt{GitHub} releases could be written individually by each user, but this creates an unnecessary technological hurdle. To make it easier for users to access \textsc{vd} via code, we offer a novel implementation of \textsc{vdd} focussing on the \texttt{R} platform, as one of the most prominent platforms for data science \citep{R-2017}.

Here we introduce a dedicated \textsc{R} package, called \texttt{datastorr} (\smurl{github.com/richfitz/datastorr}), that facilitates access to releases of any \textsc{vd} hosted on \texttt{GitHub} (Fig. \ref{fig:technology_stack}). Specifically, the  \texttt{datastorr} package:
\begin{enumerate}
  \item Constructs the shell of a second, dataset-specific \textsc{R} package, which is used to access releases from a specific repository stored on \texttt{GitHub}.
  \item Contains the main code needed to interact with the \texttt{GitHub} \textsc{api} to retrieve versions of the dataset.
\end{enumerate}
Using \texttt{datastorr}, a researcher can create and distribute a custom \textsc{R} package that facilitates access to their data with (very) minimal computational skills.

For example, \texttt{datastorr} was used to build the package \texttt{baad.data} (\smurl{github.com/traitecoevo/baad.data}), which is an interface to the Biomass and Allometry Database (\textsc{BAAD}) \citep{Falster-2015} stored at \smurl{github.com/dfalster/baad}. The R package \texttt{baad.data} consists of only a few simple functions, that were automatically generated (along with the associated help files) with \texttt{datastorr}. For a user, accessing a version of the data is a simple as typing a single line of code (Fig. \ref{fig:technology_stack}). Accessing a different version of the data involves changing only the version number. From the users perspective, the existence of the \texttt{baad.data} and \texttt{datastorr} packages makes reproducing analyses using specific versions of the data possible \citep[e.g.][]{Duursma-2016,Falster-2016}.

Using \texttt{datastorr}, dataset \emph{maintainers} can set up their own R package to deliver there versioned dataset simply by providing:
\begin{enumerate}
  \item a \texttt{GitHub} repository name (e.g., ``traitecoevo/baad.data'') where releases are stored
  \item the filename in the release that will contain the \textsc{vd}
  \item the function used to load this into \texttt{R}.
\end{enumerate}
A full tutorial explaining precisely how to set this up is available at \smurl{github.com/ropenscilabs/datastorr}.

Then as the dataset grows over time, the \emph{maintainers} update the \texttt{git} repo and create a \texttt{GitHub} release with a new version number. All the releases are simultaneously available to any user, both point-and-click and programmatically.

The dataset-specific packages created by \texttt{datastorr} are designed to be computationally efficient and also work offline. Packages created by \texttt{datastorr} contain no actual data, only the rules for fetching the data. As such, the basic package structure is quick to install and takes up virtually no space on the user's hard-drive. The package functions by fetching each data version once (the first time it is requested), and then caching these files locally for future reuse. Moreover, \emph{users} can have several versions of the same dataset on their computer and unambiguously access the different versions with one simple function call.

\section{Discussion}

% General problem. An example of solution. Key features

@@

%work on this paragraph more
We strongly suspect that the issues we have identified in this article will not come as news to most readers: datasets are constantly evolving (hopefully improving) and despite tremendous advances in data sharing and associated technologies over the last decade, there is little consensus about what to do with this fact. In some individual cases, such as \texttt{GenBank} (see \url{https://www.ncbi.nlm.nih.gov/genbank/sequenceids/}), researchers have thought about versioning and developed specific solutions. But so far these as we mentioned above, these datasets tend to be those managed by large, well-funded institutions or consortia and involve standard data types (e.g., sequences). But for an individual research group working with a specialized type of data, creating their own \texttt{GenBank} is clearly a non-starter. As a result, we argue that there is a substantial need for a easy, cheap, and flexible solution and suggest that adopting a \textsc{vdd} system may be one way forward.

% Moved from above, need to integrate better
We have created \texttt{datastorr} to make \textsc{vdd} easy to implement and manage for \texttt{R} users. And as we stated above, this has already been used to release a variety of datasets (Table \ref{tab:examples}).
% Reword next senetnce
However, we can imagine alternative methods for interacting with \texttt{GitHub} and building templates and adopting a \textsc{vdd} system.  The key is not the specific technology, but rather the concept of creating, maintaining, and distributing versioned data.

Many of the key roadblocks preventing a switch from a static to a dynamic data world were technological: in the past, it took great deal of money and expertise to set up and maintain a \textsc{vdd} system, but now there are many current platforms for storing and distributing versioned controlled data for low or no cost.  The website \texttt{GitHub} (\smurl{www.github.com}) is a commercial platform for hosting and interacting with \texttt{git} repositories. In practice, each dataset should be a separate \texttt{git} repository on \texttt{GitHub} (see Table \ref{tab:examples} for examples). Although mainly used for computer code, \texttt{GitHub} is now also being use to manage scientific data \citep{Perkel-2016}. Maintaining the version controlled data in the cloud has two main benefits: first, it provides a platform for multiple data contributors to sync their files and correspond about changes in the dataset, and second, it hosts a stream of data releases, thus acting as a central point for  the collection, curation, and distribution of the data.  The proliferation of cloud-based tools and open-source software have now reduced these roadblocks considerably.  We argue that by linking together the tools described above, researchers can now produce, curate, and distribute \textsc{vd} much more easily and cheaply than in the past.

%The key features of a \textsc{vdd} system that are already possible are a versioning system to communicate small and large changes to users, both point-and-click and programmatic access to all versions, reliable distribution to many users, and stable, organized \textsc{doi}s (see Fig. \ref{fig:technology_stack}).  Moreover, it is relatively easy to update and add new versions.  Technology is moving forward rapidly and so it is very likely that the particular tools we describe above will add features and be even easier to use.  But, nonetheless, workable systems are already possible.

@@

%Systematically archiving data is a key part of good workflow \citep{Wilkinson-2016, Piwowar-2011, Whitlock-2011}, and tools for creating and archiving digital objects have seen rapid progress in recent years.  At present, each version of the dataset can be automatically archived and assigned a \textsc{doi}, via several providers (Table \ref{tab:doi_minting}).  Here we describe the particular integration of \texttt{GitHub} and \texttt{Zenodo}.  It is useful to have the ability to either be specific---reference a particular version for the purposes of reproducibility or general--reference the entire project.  \texttt{Zenodo} currently automatically supplies both ``concept'' and ``version'' \textsc{doi}'s to the commits associated with \texttt{GitHub} releases automatically \citep[Fig. \ref{fig:semantic} and ][]{Nielsen-2017}.  This allows users to very precisely reference either generally or specifically, both of which may be appropriate in different situations.

@@

Because \texttt{git} was built for tracking changes in code, there are aspects of how it behaves that are not natural to apply to  data \citep{Perkel-2016}.  For example, because it tracks changes line-by-line rather than cell-by-cell, there are inefficiencies in how the software stores the changes.
There is rapid current work on new implementations which will be faster computationally, smaller in file size, flexible with respect to data structures, and thus allow for larger datasets to be versioned efficiently\citep{Fli, Dat}.
% Add CKAN, rdrop2 and Git-LFS, conceptually at least, and say something useful about the future

@@

\clearpage
